National demographic-disturbance relationships and local monitoring both reduce uncertainty in projections of boreal caribou population viability: tools to support monitoring decisions

### 2023-10-25 analysis/paper/paper.Rmd
chipping away at manuscript

Demographic data for caribou populations, including abundance, vital rates and population growth, are most often obtained through the tracking of caribou cows marked with telemetry collars [e.g. @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@johnson_science_2020]. Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with collars carrying  Very High Frequency (VHF) and/or Global Positioning Satellite (GPS) transmitters allowing remote detection and data retrieval [see @nbckc_nbckc_2021]. Because these collars send out a mortality signal when no longer active after a predetermined period of time, adult survival can be estimated by tracking the fate of marked individuals. See @kaplan_nonparametric_1958; @pollock_survival_1989 for original approach and @mcloughlin_declines_2003; @hervieux_widespread_2013; @eacker_web-based_2019 for examples of application to caribou. Male caribou are not frequently collared.

Collared females are also used in most estimates of recruitment. Recruitment surveys involve locating collared individuals by aircraft and then surveying the entire group of caribou to estimate the ratio of calves to cows in the group [@nbckc_nbckc_2022]. Because this estimate is based on more than just the collared sample of cows, the sample size for recruitment estimates is typically larger than for survival (see multiplier parameter $w$, cowMult below). Note that occasionally young bulls can be misidentified as cows (parameter $u$ below; refs? ), and there is also a possibility that some calves may be missed during aerial composition surveys (parameter $z$ below). We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (parameter $c$; see Supplement B and below for details). Telemetry and aerial survey data from these methods can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for monitoring population change of many caribou herds [e.g. @decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019].

@@

To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a method for simulating local population dynamics and monitoring. This method produces example population trajectories for which the true state is known, allowing us to assess the performance of our proposed method for integrating local demographic observations with national disturbance-demographic relationships in a Bayesian population model.

@@

To simulate demographic trajectories for example populations
we used a modified version of Johnson et al's [-@johnson_science_2020]
two-stage demographic model described by Dyson et al.
[-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Maximum
potential recruitment rate is adjusted for sex ratio, misidentification
biases (Supplement B) and (optionally) delayed age at first reproduction
[@eacker_web-based_2019; @decesare_estimating_2012]:
\begin{equation}
  \dot{X}_t=\frac{c\dot{R}_t/2}{1+c\dot{R}_t/2};c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)},
  (\#eq:dotX)
\end{equation}
where $w$ is a multiplier that defines the apparent number of adult
females in the composition survey as a function of the number of
collared animals, $q$ is the ratio of young bulls to adult females, $u$
is the probability of misidentifying young bulls as adult females and
vice versa, and $z$ is the probability of missing a calf.

@@

Given the values of parameters $p_0,p_k,a, and b$ used by Johnson et al. (2020) (Table A1), recruitment rate is lowest
$(0.5\dot{X}_t)$ when $\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$
at intermediate population sizes, and declines to $0.6\dot{X}_t$ as the
population reaches carrying capacity of $k=100$ times the initial
population size. The post-juvenile female population in the next year
includes both survivors and new recruits:
\begin{equation}
\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t).
  (\#eq:dotN)
\end{equation}

@@

Following Eacker et al. [-@eacker_web-based_2019], the observed number
of calves $\hat{J}_t$ is a binomially distributed function of the
observed number of adult female caribou $\hat{W}_t$ and the estimated
recruitment rate $R_t$:
\begin{equation}
\hat{J}_t \sim \text{Binomial}(\hat{W}_t,R_t).
  (\#eq:hatJEst)
\end{equation}

We model recruitment probability as a function of anthropogenic
disturbance and fire with a log link [@johnson_science_2020;
@stewart_climate-informed_2023; @dyson_existing_2022] and Gaussian
distributed random year effect [@eacker_web-based_2019]:
\begin{equation}
\log(R_t)=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t+\epsilon^R_t; \epsilon^R_t \sim \text{Normal}(0,\sigma^2_{R}).
  (\#eq:R)
\end{equation}

Estimated recruitment probability is adjusted for sex ratio,
misidentification biases (Supplement B), and age of first reproduction
[@eacker_web-based_2019; @decesare_estimating_2012] to get expected
recruits per female:
\begin{equation}
X_t=\frac{cR_t/2}{1+cR_t/2}; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
  (\#eq:X)
\end{equation}

The composition survey bias correction term $c$ depends on the apparent
number of adult females per collared animal $w$, the ratio of young
bulls to adult females $q$, the probability of misidentifying young
bulls as adult females and vice versa $u$, and the probability of
missing a calf $z$ (Supplement B). To maintain consistency with Johnson
et al. [-@johnson_science_2020] we also provide an option to only adjust
for sex ratio and misidentification biases $X_t=cR_t/2$.

In the Bayesian model, prior uncertainty about the value of the bias
correction term $c$ is approximated with a Log-normal distribution
(Supplement B). Given the apparent number of adult females per collared
animal $w$ in each monitoring scenario (Table
\@ref(tab:monitoringScenarios)), we calculate the mean and standard
deviation of 10,000 samples of $\log{c}$ assuming the ratio of young
bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult
misidentification probability $u$ varies uniformly between 0 and 0.2,
and the probability of missing a calf $z$ varies uniformly between 0 and
0.2.

Also following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the survival R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a Gaussian
distributed function of the true survival probability:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).
  (\#eq:hatSEst)
\end{equation}

As in the national demographic model, we model survival probability as an adjusted function
of buffered anthropogenic disturbance $A_t$ with a log link
[@johnson_science_2020; @stewart_climate-informed_2023;
@dyson_existing_2022] and Gaussian distributed random year effect
[@eacker_web-based_2019]:
\begin{equation}
S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).
  (\#eq:S)
\end{equation}

@@

We assume a simple two-class demographic model, with a census in late
winter of surviving females and new recruits to the post-juvenile (aged
1+) class. The number of post-juvenile females that survive from year
$t$ to the census $W_t$ is binomially distributed with survival
probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The number of
juveniles recruiting to the post-juvenile class at the census is a
Poisson distributed function of the number of surviving post-juvenile
females and the adjusted recruitment rate $X_t$:
$J_{t} \sim \text{Poisson}(X_tW_t)$. The post-juvenile female population
in the next year includes both survivors and new recruits:
$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
stable estimates of population status, we report average population
growth over a period of $y$ years, calculated as
$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
post-juvenile population size $N_0$ is 1000 animals. For reference, all
variables and parameters are defined in Table A1.

\begin{equation}
J_{t} \sim \text{Poisson}(W_tR_t/2)
  (\#eq:J)
\end{equation}

@@

TO DO: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
Ensure notation consistency amongst manuscript, UI, documentaiton and
code where appropriate.

@@

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, it is important to know whether a
population is declining or not. If uncertainty about population status
is low, additional investment in reducing uncertainty about the
population growth rate may not be necessary or warranted. Thus, in this
analysis our outcome of interest is how often the Bayesian posterior
mean correctly predicts the true population status (increasing or
decreasing). To calculate this outcome for each disturbance and
monitoring scenario (Figure \@ref(fig:figDisturbance)) we fit the
Bayesian model to 200 example population trajectories (e.g. Figure
\@ref(fig:exampleTrajectories)). The probability of correct status
assessment for that scenario is the percentage of these 200 examples in
which the Bayesian posterior mean population growth rate and the true
population growth rate are both above or below 1.

@@

To make this work more user friendly and accessible to decision makers
we built a user interface that allows for the exploration of different
monitoring and anthropogenic disturbance scenarios. We built the user
interface as an R Shiny app [@chang_shiny_2023], building off and expanding an
app published by Eacker et al. [-@eacker_web-based_2019]. The app allows
users to provide existing local data or simulate data based on national
models as shown above, define an anthropogenic disturbance scenario and
make projections. The results shown are similar to Figure
\@ref(fig:examplePredictions) and allow for comparisons to the national
model. Many parameters controlling the Bayesian model, local data
simulation and antropogenic scenario are exposed in the app with
sensible defaults set, such that novice users can get reasonable results
quickly while advanced user can customise the projections to suite their
needs. We have also made the R code needed to run the app and carry out
the above analysis available as an R package, caribouMetrics. Making our
code available as an R package ensures that it is flexible and can be
re-used or expanded upon in future projects or combined with other tools
and packages (eg SpaDES or SyncroSym). The caribouMetrics package is
available on GitHub at <https://github.com/LandSciTech/caribouMetrics>
and a package website with detailed documentation is also available
(<https://landscitech.github.io/caribouMetrics/>).

>>>>

Demographic data for caribou populations, including abundance, vital rates and population growth, are often obtained through the tracking of caribou cows marked with telemetry collars [e.g. @courtois_aerial_2003; @rettie_dynamics_1998;@hervieux_widespread_2013;@johnson_science_2020]. Caribou cows are typically captured via net-gun from aircraft in late winter, and affixed with collars carrying  Very High Frequency (VHF) and/or Global Positioning Satellite (GPS) transmitters allowing remote detection and data retrieval [see @nbckc_nbckc_2021]. Because these collars send out a mortality signal when no longer active after a predetermined period of time, adult survival can be estimated by tracking the fate of marked individuals (see @kaplan_nonparametric_1958 and @pollock_survival_1989 for original approach, and @mcloughlin_declines_2003,  @hervieux_widespread_2013 and @eacker_web-based_2019 for examples of application to caribou). Male caribou are not frequently collared.

Collared females are also used in most estimates of recruitment. Recruitment surveys involve locating collared individuals by aircraft and then surveying the entire group of caribou to estimate the ratio of calves to cows in the group [@nbckc_nbckc_2022]. Because this estimate is based on more than just the collared sample of cows, the sample size for recruitment estimates is typically larger than for survival (see multiplier parameter $w$, cowMult below). Note that occasionally young bulls can be misidentified as cows (parameters $\dot{u}$ and $u$ below; refs? ), and there is also a possibility that some calves may be missed during aerial composition surveys (parameters $\dot{z}$ and $z$ below). We acknowledge additional uncertainty and potential bias due to misidentification of sex or missing calves with a correction term derived from a simple model of the recruitment survey observation process (see Supplement B and below for details). Telemetry and aerial survey data from these methods can be combined in a simple recruitment-mortality population model [@hatter_moose_1991] that serves as the basis for monitoring population change of many caribou herds [e.g. @decesare_estimating_2012; @serrouya_comparing_2017; @eacker_web-based_2019].

@@

To investigate the effectiveness of various monitoring strategies for reducing uncertainty about population viability, and to show how this varies with landscape condition, we begin with a model for simulating example population dynamics and observations of those dynamics. We refer to this as the "national demographic model" because it is informed by national demographic-disturbance relationships. We use simulated trajectories to assess the performance of our proposed method for integrating local demographic observations with national disturbance-demographic relationships in a Bayesian population model. For reference, all variables and parameters are defined in supplemental Table A1.

@@

To simulate demographic trajectories for example populations
we used a modified version of Johnson et al's [-@johnson_science_2020]
two-stage demographic model described by Dyson et al.
[-@dyson_existing_2022]. Given a population of post-juvenile females at the beginning of year $t$, $\dot{N}_t$, the number that survive to the census, $\dot{W}_t$, is binomially
distributed with true survival probability $\dot{S}_t$:
$\dot{W}_{t} \sim \text{Binomial}(\dot{N}_t,\dot{S}_t)$. Maximum
potential recruitment rate is adjusted for sex ratio, misidentification
biases (Supplement B) and (optionally) delayed age at first reproduction
[@eacker_web-based_2019; @decesare_estimating_2012]:
\begin{equation}
  \dot{X}_t=\frac{\dot{c}\dot{R}_t/2}{1+\dot{c}\dot{R}_t/2};\dot{c}=\frac{w(1+\dot{q}\dot{u}-\dot{u})}{(w+\dot{q}\dot{u}-\dot{u})(1-\dot{z})},
  (\#eq:dotX)
\end{equation}
where $w$ is a multiplier that defines the apparent number of adult
females in the composition survey as a function of the number of
collared animals, $\dot{q}$ is the ratio of young bulls to adult females, $\dot{u}$
is the probability of misidentifying young bulls as adult females and
vice versa, and $\dot{z}$ is the probability of missing a calf. We examine sensitivity to these unknown parameters by selecting values for each simulated example population from plausible ranges: $\dot{u}$ and $\dot{v}$ vary uniformly between 0 and 0.2, and $\dot{q}$ varies uniformly between 0 and 0.6.

@@

Given the values of parameters $p_0,p_k,a,$ and $b$ used by Johnson et al. (2020) (Table A1), recruitment rate is lowest
$(0.5\dot{X}_t)$ when $\dot{N}_t=1$, approaches a maximum of $\dot{X}_t$
at intermediate population sizes, and declines to $0.6\dot{X}_t$ as the
population reaches carrying capacity of $k=100$ times the initial
population size. The post-juvenile female population in the next year
includes both survivors and new recruits:
\begin{equation}
\dot{N}_{t+1}=\text{min}(\dot{W}_t+\dot{J}_t,r_{max}\dot{N}_t).
  (\#eq:dotN)
\end{equation}

@@

Simulated local demographic and covariate data are provided as inputs to the Bayesian population model, including the observed state (dead or alive) of each collared animal in each month ($\hat{I}_{i,t,m}$), the observed number of adult females and calves in the recruitment surveys ($\hat{W}_t, \hat{J}_t$), the apparent number of adult females per collared female in the composition survey $w$, and the measures of anthropogenic disturbance and fire ($A_t, F_t$). Following Eacker et al. [-@eacker_web-based_2019], recruitment rate $R_t$ is estimated from the observed number of calves $\hat{J}_t$ and adult female caribou $\hat{W}_t$ assuming a binomial distribution:
\begin{equation}
\hat{J}_t \sim \text{Binomial}(\hat{W}_t,R_t).
  (\#eq:hatJEst)
\end{equation}

Where appropriate, the Bayesian population model and national demographic model share the same structure, and we have selected notation to highlight the similarities. The difference between equations \@ref(eq:hatJ) and \@ref(eq:hatJEst) is that in the simulation model an unadjusted apparent recruitment rate $\dot{R}_t$ is selected and used to simulate observations. In Bayesian model observations are provided, and used to estimate the unobserved recruitment rate $R_t$.

In both models recruitment probability is a function of anthropogenic
disturbance and fire with a log link [@johnson_science_2020;
@stewart_climate-informed_2023; @dyson_existing_2022]. In the Bayesian model we add a Gaussian
distributed random year effect [@eacker_web-based_2019]:
\begin{equation}
\log(R_t)=\beta^R_0+\beta^R_a A_t+\beta^R_f F_t+\epsilon^R_t; \epsilon^R_t \sim \text{Normal}(0,\sigma^2_{R}).
  (\#eq:R)
\end{equation}

In the simulation model (eq \@ref(eq:barR)), regression coefficient values ($\dot{\beta}^R_0,\dot{\beta}^R_a,\dot{\beta}^R_f$) are selected for each example population and used to calculate recruitment rate. In the Bayesian model (eq \@ref(eq:R)), unobserved regression coefficients ($\beta^R_0,\beta^R_a,\beta^R_f$) are estimated from local recruitment survey and disturbance observations.

To account for bias and additional uncertainty caused by unknown misidentification biases (Supplement B) and delayed age of age of first reproduction
[@eacker_web-based_2019; @decesare_estimating_2012] the estimated recruitment probability is adjusted to get expected recruits per female:
\begin{equation}
X_t=\frac{cR_t/2}{1+cR_t/2}; c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}.
  (\#eq:X)
\end{equation}

The composition survey bias correction term $c$ depends on the known apparent
number of adult females per collared animal $w$, the unknown ratio of young
bulls to adult females $q$, the unknown probability of misidentifying young
bulls as adult females and vice versa $u$, and the unknown probability of
missing a calf $z$ (Supplement B). Prior uncertainty about the value of thebias
correction term $c$ is approximated with a Log-normal distribution
(Supplement B). Given the apparent number of adult females per collared
animal $w$ in each monitoring scenario (Table
\@ref(tab:monitoringScenarios)), we calculate the mean and standard
deviation of 10,000 samples of $\log{c}$ assuming the ratio of young
bulls to adult females $q$ varies uniformly between 0 and 0.6, the adult
misidentification probability $u$ varies uniformly between 0 and 0.2,
and the probability of missing a calf $z$ varies uniformly between 0 and
0.2. To maintain consistency with Johnson
et al. [-@johnson_science_2020] we also provide an option to only adjust
for sex ratio and misidentification biases $X_t=cR_t/2$.

Following Eacker et al. [-@eacker_web-based_2019], Kaplan-Meier
estimates of observed adult female survival probability $\hat{S}_t$ and
associated precisions $\tau_t$ are estimated from known-fate radio
collar data using the survival R package [@therneau_modeling_2000], with
precisions for years with no observed mortality estimated using a
separate Bayesian model. Observed estimated survival is a Gaussian
distributed function of the true survival probability:
\begin{equation}
\hat{S_t} \sim \text{Normal}(S_t,1/\tau_t).
  (\#eq:hatSEst)
\end{equation}

As in the simulations (eq \@ref(eq:barS)), we model survival probability as an adjusted function of buffered anthropogenic disturbance $A_t$ with a log link
[@johnson_science_2020; @stewart_climate-informed_2023; @dyson_existing_2022]. We add a Gaussian distributed random year effect [@eacker_web-based_2019]:
\begin{equation}
S_t=\text{min}(46 \tilde{S_t}-0.5)/45,1); \log(\tilde{S_t})=\beta^S_0+\beta^S_a A_t+\epsilon^S_t; \epsilon^S_t \sim \text{Normal}(0,\sigma^2_{S}).
  (\#eq:S)
\end{equation}

@@

As in the simulations, we assume a simple two-class demographic model, with a census in late
winter of surviving females and new recruits to the post-juvenile (aged
1+) class. The number of post-juvenile females that survive from year
$t$ to the census $W_t$ is binomially distributed with survival
probability $S_t$: $W_{t} \sim \text{Binomial}(N_t,S_t)$. The number of
juveniles recruiting to the post-juvenile class at the census is a
Poisson distributed function of the number of surviving post-juvenile
females and the adjusted recruitment rate $X_t$:
$J_{t} \sim \text{Poisson}(X_tW_t)$. The post-juvenile female population
in the next year includes both survivors and new recruits:
$N_{t+1}=W_t+J_t$. Annual population growth rate $\lambda_t$ is set to 0
when $N_t$ is 0, and is otherwise $\lambda_t=N_{t+1}/N_t$. For more
stable estimates of population status, we report average population
growth over a period of $y=3$ years, calculated as
$\bar{\lambda}_{y,t}=\sum_{x=t-y+1}^{x=t}\lambda_x$. The initial
post-juvenile population size $N_0$ is 1000 animals. For reference, all
variables and parameters are defined in supplemental Table A1.

@@

<!--
Gritty details for the record: \* note distribution of random effect parameters are truncated to
ensure R and S remain within reasonable bounds (not guaranteed given log
link). Not yet documented here. \* note Eacker survival method gives
biased results in years when \<12 months of survival data is available.
\* Note I am dropping survival information for animals that are not
alive at the beginning of a year to deal with censoring issues. There is
probably some more elegant way to deal with this, but it won't make any
difference as long as the collar deployment month is set to 1. \*
consider & comment on alternative choices made in this example
<https://poissonconsulting.shinyapps.io/bboushiny/>. Follow Eacker
example in all particulars, change some aspects, or do comparisons? \*
-->

@@

In order to assess whether a monitoring program is likely sufficient to
inform decisions, it is important to select outcome measures that are
relevant to decision making [@gregory_structured_2012]. For boreal
caribou and other endangered species, a relevant outcome is whether a
population is declining or not. If uncertainty about population status
is low, additional investment in reducing uncertainty about the
population growth rate may not be necessary or warranted. Thus, in this
analysis our outcome of interest is how often the Bayesian posterior
mean correctly predicts the true population status (increasing or
decreasing). To calculate this outcome for each disturbance and
monitoring scenario (Figure \@ref(fig:figDisturbance)) we fit the
Bayesian model to 200 example population trajectories (e.g. Figure
\@ref(fig:exampleTrajectories)). The probability of correct status
assessment for that scenario is the percentage of these 200 examples in
which the Bayesian posterior mean population growth rate and the true
population growth rate are both above or below 1.

@@

To make this work more user friendly and accessible we built an R Shiny [@chang_shiny_2023] user interface that allows for exploration of monitoring and anthropogenic disturbance scenarios. Where appropropriate, we borrowed and built on code from an app developed by Eacker et al. [-@eacker_web-based_2019]. Our app allows
users to provide existing local data or simulate data based on national
models as shown above, define an anthropogenic disturbance scenario and
make projections. The results shown are similar to Figure
\@ref(fig:examplePredictions) and allow for comparisons to the national
model. The app allows advanced users to adjust most model parameters (supplemental Table A1). Default parameter values are set to allow novice users to get reasonable results quickly. Most of the code needed to run the app is in an R package, caribouMetrics, that is designed to allow the national demographic simulation model and the Bayesian population model to be re-used in a variety of contexts (e.g. Frances SpaDES paper, Dyson preprint, syncrosim reference). The caribouMetrics package is available on GitHub at <https://github.com/LandSciTech/caribouMetrics>. Find documentation and vignettes on the package website
<https://landscitech.github.io/caribouMetrics/>. The shiny app is available at <https://landscitech.github.io/BayesianCaribouDemographicProjection/>, and code required to run analyses and produce figures in the paper is available at <https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>.

### 2023-10-25 analysis/paper/supplement.Rmd
chipping away at manuscript

|**Symbol**|**Units**|**Description**|
|:-----|-----|:-----------|
|**Indices & constants**|
|$t$|year|Year.|
|$m$|month|Month.|
|$i$|ID|Individual animal identifier.|
|**Observed**|
|$A_t$|%|Buffered anthropogenic disturbance.|
|$F_t$|%|Fire disturbance.| 
|$\hat{J}_t$|animals|Calves observed in calf:cow survey.|
|$\hat{W}_t$|animals|Cows observed in calf:cow survey.|
|$\hat{T}_t$|animals|Number of collared cows.|
|$\hat{S}_t$|probability|Kaplan-Meier estimate of observed adult female survival.|
|$\tau_t$||Precision of Kaplan-Meier survival estimate.|
|$\hat{I}_{i,t,m}$|0 or 1|Indicator of whether animal $i$ is dead or alive.|
|**Estimated state**|
|$R_t$|rate|Recruitment (calves per cow).|
|$S_t$|probability|Adult female survival.|
|$N_t$|animals|Adult female population size.|
|$\bar{\lambda}_{y,t}$|rate|Average population growth over $y$ years.|
|$X_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|
|$W_t$|animals|Surviving adult females before recruitment.|
|$J_t$|animals|New recruits to adult age class.|
|$\lambda_t$|rate|Annual population growth.|
|**Estimated Parameters**|
|$\beta^R_0$||Recruitment intercept.|
|$\beta^R_a$||Anthropogenic disturbance slope for recruitment.|
|$\beta^R_f$||Fire slope for recruitment.|
|$\sigma^2_{R}$||Variance in random effect of year on recruitment.|
|$\beta^S_0$||Survival intercept.|
|$\beta^S_a$||Anthropogenic disturbance slope for survival.|
|$\sigma^2_{S}$||Variance in random effect of year on survival.|
|**True state**|
|$\dot{R}_t$|rate|Recruitment (calves per cow).|
|$\dot{S}_t$|probability|Adult female survival.|
|$\dot{N}_t$|animals|Adult female population size.|
|$\dot{X}_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|
|$\dot{W}_t$|animals|Surviving adult females before recruitment.|
|$\dot{J}_t$|animals|New recruits to adult age class.|
|**True demographic parameters**|
|$N_0$|animals|Initial adult population size.|
|$\bar{R}_t$|rate|Expected recruitment (calves per cow).|
|$\bar{S}_t$|probability|Expected adult female survival.|
|$\dot{\beta}^R_0$||Recruitment intercept.|
|$\dot{\beta}^R_a$||Anthropogenic disturbance slope for recruitment.|
|$\dot{\beta}^R_f$||Fire slope for recruitment.|
|$\dot{\beta}^S_0$||Survival intercept.|
|$\dot{\beta}^S_a$||Anthropogenic disturbance slope for survival.|
|$p_0=1$||Maximum recruitment multiplier.|
|$p_k=0.6$||Recruitment multiplier at carrying capacity.|
|$k=100$||Carrying capacity multiplier.|
|$b=4$||Density dependence shape parameter.|
|$a=1$||Allee effect parameter.|
|$r_{max}=1.3$||Maximum population growth rate.|
|$\nu_R=0.46$||Coefficient of variation in $\dot{R}_t$ among years.|
|$\nu_S=0.08696$||Coefficient of variation in $\dot{S}_t$ among years.|
|$l_R=0$||Minimum value for $\dot{R}_t$.|
|$l_S=0$||Minimum value for $\dot{S}_t$.|
|$h_R=0$||Maximum value for $\dot{R}_t$.|
|$h_S=0$||Maximum value for $\dot{S}_t$.|
|**Observation model parameters**|
|$d=1,2,4,8,16, \text{ or } 24$|years|Monitoring duration.|
|$n=15,30, \text{ or } 60$|animals|Target number of collared females.|
|$o=1 \text{ or } 4$|years|Years between collar deployments.|
|$y=3$|years|Assessment period for population growth rate.|
|$w=3,6,\text{ or } 9$|ratio|The apparent number of adult females per collared animal in composition survey.|
|$q=\text{uniform}(0,0.6)$|ratio|True ratio of young bulls to adult females in composition survey.|
|$u=\text{uniform}(0,0.2)$|probability|Probability of misidentifying young bulls as adult females and vice versa in composition survey.|
|$z=\text{uniform}(0,0.2)$|probability|Probability of missing calves in composition survey.|
|$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}$||Bias correction term (Appendix X).|

>>>>

|**Symbol**|**Units**|**Description**|
|:-----|-----|:-----------|
|**Indices & constants**|
|$t$|year|Year.|
|$m$|month|Month.|
|$i$|ID|Individual animal identifier.|
|**Observed**|
|$A_t$|%|Buffered anthropogenic disturbance.|
|$F_t$|%|Fire disturbance.| 
|$\hat{J}_t$|animals|Calves observed in calf:cow survey.|
|$\hat{W}_t$|animals|Cows observed in calf:cow survey.|
|$\hat{T}_t$|animals|Number of collared cows.|
|$\hat{S}_t$|probability|Kaplan-Meier estimate of observed adult female survival.|
|$\tau_t$||Precision of Kaplan-Meier survival estimate.|
|$\hat{I}_{i,t,m}$|0 or 1|Indicator of whether animal $i$ is dead or alive.|
|**Estimated state**|
|$R_t$|rate|Recruitment (calves per cow).|
|$S_t$|probability|Adult female survival.|
|$N_t$|animals|Adult female population size.|
|$\bar{\lambda}_{y,t}$|rate|Average population growth over $y$ years.|
|$X_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|
|$W_t$|animals|Surviving adult females before recruitment.|
|$J_t$|animals|New recruits to adult age class.|
|$\lambda_t$|rate|Annual population growth.|
|**Estimated Parameters**|
|$\beta^R_0$||Recruitment intercept.|
|$\beta^R_a$||Anthropogenic disturbance slope for recruitment.|
|$\beta^R_f$||Fire slope for recruitment.|
|$\sigma^2_{R}$||Variance in random effect of year on recruitment.|
|$\beta^S_0$||Survival intercept.|
|$\beta^S_a$||Anthropogenic disturbance slope for survival.|
|$\sigma^2_{S}$||Variance in random effect of year on survival.|
|**True state**|
|$\dot{R}_t$|rate|Recruitment (calves per cow).|
|$\dot{S}_t$|probability|Adult female survival.|
|$\dot{N}_t$|animals|Adult female population size.|
|$\dot{X}_t$|rate|Recruitment adjusted for sex ratio and (optionally) delayed age of first reproduction.|
|$\dot{W}_t$|animals|Surviving adult females before recruitment.|
|$\dot{J}_t$|animals|New recruits to adult age class.|
|**True demographic parameters**|
|$N_0$|animals|Initial adult population size.|
|$\bar{R}_t$|rate|Expected recruitment (calves per cow).|
|$\bar{S}_t$|probability|Expected adult female survival.|
|$\dot{\beta}^R_0$||Recruitment intercept.|
|$\dot{\beta}^R_a$||Anthropogenic disturbance slope for recruitment.|
|$\dot{\beta}^R_f$||Fire slope for recruitment.|
|$\dot{\beta}^S_0$||Survival intercept.|
|$\dot{\beta}^S_a$||Anthropogenic disturbance slope for survival.|
|$p_0=1$||Maximum recruitment multiplier.|
|$p_k=0.6$||Recruitment multiplier at carrying capacity.|
|$k=100$||Carrying capacity multiplier.|
|$b=4$||Density dependence shape parameter.|
|$a=1$||Allee effect parameter.|
|$r_{max}=1.3$||Maximum population growth rate.|
|$\nu_R=0.46$||Coefficient of variation in $\dot{R}_t$ among years.|
|$\nu_S=0.08696$||Coefficient of variation in $\dot{S}_t$ among years.|
|$l_R=0$||Minimum value for $\dot{R}_t$.|
|$l_S=0$||Minimum value for $\dot{S}_t$.|
|$h_R=0$||Maximum value for $\dot{R}_t$.|
|$h_S=0$||Maximum value for $\dot{S}_t$.|
|**Observation model parameters**|
|$d=1,2,4,8,16, \text{ or } 24$|years|Monitoring duration.|
|$n=15,30, \text{ or } 60$|animals|Target number of collared females.|
|$o=1 \text{ or } 4$|years|Years between collar deployments.|
|$y=3$|years|Assessment period for population growth rate.|
|$w=3,6,\text{ or } 9$|ratio|The apparent number of adult females per collared animal in composition survey.|
|$\dot{q}=\text{uniform}(0,0.6)$|ratio|True ratio of young bulls to adult females in composition survey.|
|$\dot{u}=\text{uniform}(0,0.2)$|probability|True probability of misidentifying young bulls as adult females and vice versa in composition survey.|
|$\dot{z}=\text{uniform}(0,0.2)$|probability|True probability of missing calves in composition survey.|
|$\dot{c}$||True bias correction term in simulation model (Supplement B).|
|$q=\text{uniform}(0,0.6)$|ratio|Unknown ratio of young bulls to adult females in composition survey.|
|$u=\text{uniform}(0,0.2)$|probability|Unknown probability of misidentifying young bulls as adult females and vice versa in composition survey.|
|$z=\text{uniform}(0,0.2)$|probability|Unknown probability of missing calves in composition survey.|
|$c$||Unknown bias correction term in Bayesian model (Supplement B).|
